---
title: "Framingham Heart Study Data Analysis - Assignment 4"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 10, fig.height = 6)
```

# Introduction

This analysis examines the Framingham Heart Study dataset, focusing on cardiovascular disease risk factors and outcomes. The Framingham Heart Study is one of the longest-running epidemiological studies, beginning in 1948.

## Load Required Libraries

```{r libraries}
library(readxl)    # For reading Excel files
library(dplyr)     # Data manipulation
library(ggplot2)   # Visualization
library(corrplot)  # Correlation plots
library(tidyr)     # Data tidying
library(knitr)     # For nice tables
```

## Load the Clean Dataset

```{r load-data}
# Load the prepared dataset
mydata <- readRDS("FHS_assign4.rds")

# Display dimensions
cat("Dataset dimensions:", dim(mydata), "\n")
cat("Rows:", nrow(mydata), "| Columns:", ncol(mydata), "\n")
```

---

# Section 1: Data Overview

## Dataset Structure

```{r data-structure}
# Display structure
str(mydata)
```

## Summary Statistics

```{r summary-stats}
# Summary of all variables
summary(mydata)
```

## Variable Names

```{r variable-names}
# List all variable names
cat("Variables in the dataset:\n")
print(names(mydata))
```

---

# Section 2: Descriptive Statistics

## Demographic Variables

```{r demographics}
# Age distribution
if("AGE1" %in% names(mydata)) {
  cat("Age Statistics:\n")
  cat("  Mean:", mean(mydata$AGE1, na.rm = TRUE), "\n")
  cat("  Median:", median(mydata$AGE1, na.rm = TRUE), "\n")
  cat("  Range:", range(mydata$AGE1, na.rm = TRUE), "\n")
}

# Sex distribution
if("SEX1" %in% names(mydata)) {
  cat("\nSex Distribution:\n")
  print(table(mydata$SEX1))
  cat("  Proportion:\n")
  print(prop.table(table(mydata$SEX1)))
}
```

## Clinical Variables

```{r clinical-vars}
# Blood pressure
if("SYSBP1" %in% names(mydata) & "DIABP1" %in% names(mydata)) {
  cat("Blood Pressure Statistics:\n")
  cat("  Systolic BP - Mean:", mean(mydata$SYSBP1, na.rm = TRUE), 
      "| SD:", sd(mydata$SYSBP1, na.rm = TRUE), "\n")
  cat("  Diastolic BP - Mean:", mean(mydata$DIABP1, na.rm = TRUE), 
      "| SD:", sd(mydata$DIABP1, na.rm = TRUE), "\n")
}

# BMI
if("BMI1" %in% names(mydata)) {
  cat("\nBMI Statistics:\n")
  cat("  Mean:", mean(mydata$BMI1, na.rm = TRUE), 
      "| SD:", sd(mydata$BMI1, na.rm = TRUE), "\n")
  cat("  Range:", range(mydata$BMI1, na.rm = TRUE), "\n")
}
```

---

# Section 3: Data Visualization

## Age Distribution

```{r age-histogram}
if("AGE1" %in% names(mydata)) {
  ggplot(mydata, aes(x = AGE1)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    labs(title = "Distribution of Age",
         x = "Age (years)",
         y = "Frequency") +
    theme_minimal()
}
```

## Blood Pressure Distribution

```{r bp-plots}
if("SYSBP1" %in% names(mydata) & "DIABP1" %in% names(mydata)) {
  library(gridExtra)
  
  p1 <- ggplot(mydata, aes(x = SYSBP1)) +
    geom_histogram(bins = 30, fill = "coral", color = "white") +
    labs(title = "Systolic Blood Pressure", x = "SBP (mmHg)", y = "Count") +
    theme_minimal()
  
  p2 <- ggplot(mydata, aes(x = DIABP1)) +
    geom_histogram(bins = 30, fill = "lightblue", color = "white") +
    labs(title = "Diastolic Blood Pressure", x = "DBP (mmHg)", y = "Count") +
    theme_minimal()
  
  grid.arrange(p1, p2, ncol = 2)
}
```

## BMI Distribution by Sex

```{r bmi-by-sex}
if("BMI1" %in% names(mydata) & "SEX1" %in% names(mydata)) {
  ggplot(mydata, aes(x = factor(SEX1), y = BMI1, fill = factor(SEX1))) +
    geom_boxplot() +
    labs(title = "BMI Distribution by Sex",
         x = "Sex (1=Male, 2=Female)",
         y = "BMI (kg/m²)") +
    theme_minimal() +
    theme(legend.position = "none")
}
```

---

# Section 4: Correlation Analysis

## Select Numeric Variables for Correlation

```{r select-numeric}
# Select only numeric variables (exclude binary indicators)
numeric_vars <- mydata %>%
  select(where(is.numeric)) %>%
  select(-starts_with("d_"))  # Exclude binary coded variables

cat("Numeric variables for correlation:\n")
print(names(numeric_vars))
```

## Correlation Matrix

```{r correlation-matrix}
# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Display correlation matrix
kable(round(cor_matrix, 3), caption = "Correlation Matrix")
```

## Correlation Heatmap

```{r correlation-plot, fig.height=8}
corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         tl.cex = 0.8,
         tl.col = "black",
         addCoef.col = "black",
         number.cex = 0.6,
         title = "Correlation Matrix of Clinical Variables",
         mar = c(0,0,2,0))
```

---

# Section 5: Outcome Variable Analysis

## CVD/CHD Outcomes

```{r outcome-analysis}
# Check for cardiovascular disease outcomes
outcome_vars <- grep("^d_", names(mydata), value = TRUE)

if(length(outcome_vars) > 0) {
  cat("Binary outcome variables found:\n")
  for(var in outcome_vars) {
    cat("\n", var, ":\n")
    print(table(mydata[[var]]))
    cat("  Prevalence:", mean(mydata[[var]] == 1, na.rm = TRUE) * 100, "%\n")
  }
}
```

## Outcome by Age Group

```{r outcome-by-age}
if("AGE1" %in% names(mydata) & "d_anychd" %in% names(mydata)) {
  # Create age groups
  mydata$age_group <- cut(mydata$AGE1, 
                          breaks = c(0, 40, 50, 60, 70, 100),
                          labels = c("<40", "40-50", "50-60", "60-70", "70+"))
  
  # CHD by age group
  chd_by_age <- mydata %>%
    group_by(age_group) %>%
    summarise(
      n = n(),
      chd_cases = sum(d_anychd == 1, na.rm = TRUE),
      prevalence = mean(d_anychd == 1, na.rm = TRUE) * 100
    )
  
  print(kable(chd_by_age, digits = 2, 
              caption = "CHD Prevalence by Age Group"))
  
  # Visualize
  ggplot(chd_by_age, aes(x = age_group, y = prevalence)) +
    geom_bar(stat = "identity", fill = "darkred") +
    labs(title = "CHD Prevalence by Age Group",
         x = "Age Group",
         y = "Prevalence (%)") +
    theme_minimal()
}
```

---

# Section 6: Risk Factor Analysis

## Blood Pressure and CVD

```{r bp-cvd}
if("SYSBP1" %in% names(mydata) & "d_cvd" %in% names(mydata)) {
  ggplot(mydata, aes(x = factor(d_cvd), y = SYSBP1, fill = factor(d_cvd))) +
    geom_boxplot() +
    labs(title = "Systolic BP by CVD Status",
         x = "CVD (0=No, 1=Yes)",
         y = "Systolic BP (mmHg)") +
    theme_minimal() +
    theme(legend.position = "none")
}
```

## BMI and Cardiovascular Outcomes

```{r bmi-outcomes}
if("BMI1" %in% names(mydata) & any(grepl("^d_", names(mydata)))) {
  # BMI categories
  mydata$bmi_category <- cut(mydata$BMI1,
                              breaks = c(0, 18.5, 25, 30, 100),
                              labels = c("Underweight", "Normal", 
                                       "Overweight", "Obese"))
  
  # Check if we have any CVD outcome
  outcome_var <- grep("^d_(cvd|anychd)", names(mydata), value = TRUE)[1]
  
  if(!is.null(outcome_var)) {
    bmi_outcome <- mydata %>%
      group_by(bmi_category) %>%
      summarise(
        n = n(),
        cases = sum(get(outcome_var) == 1, na.rm = TRUE),
        prevalence = mean(get(outcome_var) == 1, na.rm = TRUE) * 100
      )
    
    print(kable(bmi_outcome, digits = 2,
                caption = paste("Outcome by BMI Category")))
  }
}
```

---

# Section 7: Statistical Tests

## T-test: Age by CVD Status

```{r t-test-age}
if("AGE1" %in% names(mydata) & "d_cvd" %in% names(mydata)) {
  # T-test
  t_result <- t.test(AGE1 ~ d_cvd, data = mydata)
  
  cat("T-test: Age by CVD Status\n")
  cat("  Mean age (No CVD):", mean(mydata$AGE1[mydata$d_cvd == 0], na.rm = TRUE), "\n")
  cat("  Mean age (CVD):", mean(mydata$AGE1[mydata$d_cvd == 1], na.rm = TRUE), "\n")
  cat("  p-value:", t_result$p.value, "\n")
  
  if(t_result$p.value < 0.05) {
    cat("  Result: Significant difference (p < 0.05)\n")
  } else {
    cat("  Result: No significant difference (p >= 0.05)\n")
  }
}
```

## Chi-Square: Sex and Hypertension

```{r chi-square}
if("SEX1" %in% names(mydata) & "d_hyperten" %in% names(mydata)) {
  # Contingency table
  cont_table <- table(mydata$SEX1, mydata$d_hyperten)
  print(cont_table)
  
  # Chi-square test
  chi_result <- chisq.test(cont_table)
  
  cat("\nChi-Square Test: Sex vs Hypertension\n")
  cat("  Chi-square statistic:", chi_result$statistic, "\n")
  cat("  p-value:", chi_result$p.value, "\n")
}
```

---

# Section 8: Simple Linear Regression

## Predicting Systolic BP from Age

```{r regression-age-sbp}
if("AGE1" %in% names(mydata) & "SYSBP1" %in% names(mydata)) {
  # Fit model
  model1 <- lm(SYSBP1 ~ AGE1, data = mydata)
  
  # Display results
  cat("Linear Regression: SBP ~ Age\n")
  print(summary(model1))
  
  # Visualize
  ggplot(mydata, aes(x = AGE1, y = SYSBP1)) +
    geom_point(alpha = 0.3, color = "steelblue") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = "Systolic BP vs Age",
         x = "Age (years)",
         y = "Systolic BP (mmHg)") +
    theme_minimal()
}
```

---

# Section 9: Multiple Regression

## Predicting Systolic BP from Multiple Factors

```{r multiple-regression}
if(all(c("SYSBP1", "AGE1", "BMI1", "SEX1") %in% names(mydata))) {
  # Fit multiple regression model
  model2 <- lm(SYSBP1 ~ AGE1 + BMI1 + SEX1, data = mydata)
  
  cat("Multiple Regression: SBP ~ Age + BMI + Sex\n")
  print(summary(model2))
  
  # Confidence intervals
  cat("\nConfidence Intervals:\n")
  print(confint(model2))
}
```

---

# Section 10: Logistic Regression (if applicable)

## Predicting CVD Risk

```{r logistic-regression}
if(all(c("d_cvd", "AGE1", "SYSBP1", "BMI1") %in% names(mydata))) {
  # Fit logistic regression
  logit_model <- glm(d_cvd ~ AGE1 + SYSBP1 + BMI1, 
                     data = mydata, 
                     family = binomial(link = "logit"))
  
  cat("Logistic Regression: CVD ~ Age + SBP + BMI\n")
  print(summary(logit_model))
  
  # Odds ratios
  cat("\nOdds Ratios:\n")
  print(exp(coef(logit_model)))
  
  # 95% CI for odds ratios
  cat("\n95% CI for Odds Ratios:\n")
  print(exp(confint(logit_model)))
}
```

---

# Conclusions

## Key Findings

1. **Dataset Characteristics**: [Summarize your cleaned dataset]

2. **Risk Factors**: [Discuss main risk factors observed]

3. **Statistical Relationships**: [Summarize significant relationships found]

4. **Model Performance**: [Discuss regression/prediction results]

## Limitations

- [List any data limitations]
- [Discuss missing data impact]
- [Note any analytical constraints]

## Future Directions

- [Suggest additional analyses]
- [Propose model improvements]

---

# Session Information

```{r session-info}
sessionInfo()
```
